Does stomatal patterning in amphistomatous leaves minimize the CO\(_2\) diffusion path length with leaves of \emph{Arabidopsis thaliana}?

### 2023-10-18 ms/ms.tex
revised discussion

\hypersetup{
  pdftitle={Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous Arabidopsis thaliana Incurs No Photosynthetic Advantage},
  pdfkeywords={amphistomy, Arabidopsis thaliana, CO\(_2\)
diffusion, finite element method, optimality, photosynthesis, stomata},
  breaklinks=true,
  bookmarks=true,
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

% Create labels for Addresses if the are given in Elsevier format
   \MyNewLabel{ADDRTXT}{Colgate University}{%
  13 Oak Drive, Hamilton, NY 13346%
 }
   \MyNewLabel{ADDRTXT}{University of Hawaii at Manoa}{%
  2500 Campus Rd, Honolulu, HI 96822%
 }

% Create labels for Footnotes if they are given in Elsevier format
\MyNewLabel{CORRESPTXT}{1}{Equal contribution}
\MyNewLabel{CORRESPTXT}{2}{Current email address:
\href{mailto:cat@example.com}{cat@example.com}}

% Pandoc header-include feature
\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{lineno}
\linenumbers
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}
\usepackage{caption}
% Pandoc header-include feature
\usepackage{booktabs}

@@

\title[]{Spatial Coordination Of Stomatal Patterning Between Leaf
Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No
Photosynthetic Advantage}

@@

\author[%
\addrlblbycode{Colgate University}%
,\refstepcounter{correspcnt}\setcounter{thisauthcorresp}{\value{correspcnt}}\fnsymbol{thisauthcorresp}%
%
%
]{Jacob L. Watts}

\addrbycode{Colgate University}

\corresp[\fnsymbol{thisauthcorresp}]{Corresponding author. \href{mailto:jwatts@colgate.edu}{\nolinkurl{jwatts@colgate.edu}}}

\author[%
\addrlblbycode{ETH}%
%
%
%
]{Graham Dow}

\addrbycode{ETH}

\author[%
\addrlblbycode{University of California Davis}%
%
%
,\corresplblbycode{1}%
]{Thomas N. Buckley}

\addrbycode{University of California Davis}

\correspbycode{1}

\author[%
\addrlblbycode{University of Hawaii at Manoa}%
%
%
,\corresplblbycode{1}%
]{Christopher D. Muir}

\addrbycode{University of Hawaii at Manoa}

\correspbycode{1}

@@

\abstract{
This is the abstract. It consists of two paragraphs.}

@@

Many successful predictions about stomata and other leaf traits can be
made by hypothesizing that natural selection should optimizes CO\(_2\)
gain per unit of water loss
\citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
However, stomatal anatomy may be partially constrained by physical and
developments limits on phenotypic expression
\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
Sometimes optimization leads to similar phenotypes across many disparate
species. For example, almost all stomata follow the one cell spacing
rule to maintain proper stomatal functioning
\citep{geisler_oriented_2000, dow_physiological_2014}; however some
species (notably in \emph{Begonia}) appear to benefit from overlapping
vapor shells caused by stomatal clustering
\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
Stomatal traits also vary adaptively in different environments. Stomatal
density positively co-varies with irradiance during leaf development and
negatively co-varies with CO\(_2\) concentration
\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
consistent with optimality predictions. Stomatal size is jointly
controlled by genome size, light, and stomatal density
\citep{jordan_environmental_2015}. Size positively co-varies with genome
size \citep{roddy_scaling_2020} and negatively co-varies with stomatal
density \citep{camargo_density_2011}. Total stomatal area (size
\(\times\) density) is optimized for operational conductance
(\(gs_\text{s,op}\)) rather than maximum conductance
(\(gs_\text{s,max}\)) such that stomatal apertures are most responsive
to changes in the environment at their operational aperture
\citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
can compensate for maladaptive stomatal densities to an extent
\citep{bussis_stomatal_2006}, but stomatal density and size ultimately
determine a leaf's theoretical \(gs_\text{s,max}\)
\citep{sack_developmental_2016}, which is proportional to
\(gs_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
stomatal densities lead to irregular and insufficient CO\(_2\) supply
and reduced photosynthetic efficiency in areas far from stomata
\citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
stomatal densities can reduce water use efficiency (WUE)
\citep{bussis_stomatal_2006} and incur excessive metabolic costs
\citep{deans_optimization_2020}. In most species, stomata occur on the
abaxial (usually lower) leaf surface; but amphistomy, the occurrence of
stomata on both abaxial and adaxial leaf surfaces, is also prevalent in
high light environments with constant or intermittent access to
sufficient water
\citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance
\citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
Historically, stomatal patterning in dicot angiosperms was thought to be
random with an exclusionary distance surrounding each stomate
\citep{sachs_developmental_1974}; however, the developmental controls of
stomatal patterning are poorly understood and likely more complex than
random development along the leaf surface.
\citet{croxdale_stomatal_2000}{]} reviews three developmental theories
which attempt to explain stomatal patterning in angiosperms: inhibition,
cell lineage, and cell cycle, ultimately arguing for a cell cycle based
control of stomatal patterning.

@@

We hypothesized that natural selection will favor stomatal patterning
and distribution to minimize the diffusion path length. In
amphistomatous leaves, this would be accomplished by 1) a uniform
distribution of stomata on both abaxial and adaxial leaf surfaces and 2)
coordinated stomatal spacing on each surface that offsets the position
of stomata (\autoref{fig:ideal-amphi-grid}). Coordination between leaf
surfaces is defined in this study as the occurrence of stomata in areas
farthest from stomata on the opposite leaf surface. Additionally,
because CO\(_2\) is more limiting for photosynthesis under high light,
we hypothesize that in high light 3) there should be more stomata, and
4) stomata should be more uniformly distributed than in low light.
Finally, as stomatal densities are selected for optimal operational
aperture, we hypothesize that 5) stomatal length will be positively
correlated with the area of the leaf surface to which it is closest. We
refer to this as the `stomatal zone', the leaf area surrounding a focal
stomate closest to that stomate and therefore the zone it supplies with
CO\(_2\)). This way, each stomate can be optimally sized relative to the
mesophyll volume it supplies.

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/ideal-amphi-grid.pdf}
\caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
\label{fig:ideal-amphi-grid}
\end{figure}

@@

\begin{table}[ht]
\centering
\begin{tabular}{ll}
  \hline
trait & relationship \\ 
  \hline
leaf thickness & + \\ 
  stomatal density & - \\ 
  leaf porosity & - \\ 
  lat.-vert. diffusion ratio & - \\ 
  temperature & - \\ 
  pressure & - \\ 
  Rubisco concentration & + \\ 
  light & + \\ 
   \hline
\end{tabular}
\caption{A summary of the hypothesized relationships between leaf traits and environmental conditions and photosynthetic advantage of stomatal spatial coordination in amphistomatous leaves.} 
\label{tab:hypotheses}
\end{table}

@@

{[}CDM: Graham Dow provided these images. We'll need to add him as a
co-author and ask him to write methods on image acquisition.{]}

@@

For each sample image, we also simulated \(10^3\) synthetic data with
\(n\) stomata ideally dispersed in an equilateral triangular grid. For
these grids, we integrated over plausible stomatal densities and then
conditioned on stomatal grids with exactly \(n\) stomata. The simulated
stomatal count was drawn from a Poisson distribution with the mean
parameter \(\lambda\) drawn from a Gamma distribution with shape \(n\)
and scale 1 \(\lambda \sim \Gamma(n, 1)\). \(\Gamma(n, 1)\) is the
posterior distribution of \(\lambda\) with a flat prior distribution.
This allows us to integrate over uncertainty in the stomatal density
from the sample image.

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/model_summary.pdf}
\caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis. The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane. $I_0$: PPFD incident on the leaf surface; $\varphi_\text{pal}$: Fraction of intercellular airspace (aka porosity), palisade; $T_\text{leaf}$: Leaf thickness; $U$: Interstomatal distance.}
\label{fig:model_summary}
\end{figure}

@@

Stomata cost resources to maintain \citep{deans_optimization_2020} and
expose leaves to risks such as hydraulic failure
\citep{wang_theoretical_2020} or infection by plant pathogens
\citep{melotto_stomatal_2017}. A theoretical, optimized plant would
minimize stomatal density while also allowing competitive gas exchange
rates for its environment so as to maximize C assimilation per unit
investment in stomata. Natural selection operates within developmental
and physical constraints to drive each plant species toward its
theoretical optimum. This study provides evidence that stomata in
\emph{Arabidopsis thaliana}, the model angiosperm, are non-randomly
distributed, favoring dispersion over clustering
(\autoref{fig:single-surface}). However, stomata are not ideally
dispersed in an equilateral triangular grid as would be optimal to
minimize CO\(_2\) diffusion path length and standardize the area
supplied by each stomate (\autoref{fig:tessellation-example}).
Additionally, when grown in high light environments, \emph{A. thaliana}
exhibited increased stomatal density rather than increased stomatal
dispersion (\autoref{fig:density}), which suggests that natural
selection has acted more strongly on developmental pathways that
modulate stomatal density than those that control stomatal dispersion.
In other words, plants optimize gas exchange by adding more stomata
rather than dispersing them more evenly across the leaf surface. This
study also demonstrates that stomata that supply larger leaf areas with
CO\(_2\) tend to be larger (\autoref{fig:length-area}). These results
could suggest that 1) the added energetic and hydraulic cost of
non-ideally dispersed stomata is negligible and therefore not acted on
by natural selection; 2) no developmental pathway exists to ensure the
ideal placement of stomata on the leaf; or 3) the regulation of stomatal
size limits the cost incurred by non-ideal stomatal dispersion.

In high light environments, amphistomy is favorable as high light
photosynthesis is limited by CO\(_2\) and amphistomy halves diffusion
path length and boundary layer resistance, thereby reducing CO\(_2\)
limitation - increasing theoretical \(A_\text{max}\). An optimal
amphistomatous leaf has offset stomata such that stomata are more likely
to appear on one leaf surface if there is not a stomata directly
opposite it on the other surface as shown in
\autoref{fig:ideal-amphi-grid}. However, our results show that leaf
surfaces are not coordinated but are independent, regardless of light
(\autoref{fig:dual-surface}). Additionally, gas exchange models show
little photosynthetic efficiency gain from abaxial-adaxial stomatal
coordination compared to anticoordination (INSERT FIG FROM MODELING). We
posit that this marginal gain is not sufficient to be acted upon
strongly by natural selection. Thus, amphistomatous plants do not
exhibit abaxial-adaxial stomatal coordination for there is little
selective advantage of it.

Our study corroborates previous studies which demonstrate that stomata
are non-randomly distributed along the leaf surface as a result of
developmental mechanisms such as spatially biased arrest of stomatal
initials \citep{boetsch_arrest_1995}, oriented asymmetric cell division
\citep{geisler_oriented_2000}, and cell cycle controls
\citep{croxdale_stomatal_2000}. We do not investigate the potential
developmental pathways that influence stomatal dispersion in this study;
however, they are important to consider as these pathways could limit
plants from reaching the theoretical peak in the adaptive landscape:
uniform stomatal dispersion. Instead, as this study suggests, plants may
simply compensate with higher stomatal density and by fitting stomatal
size to the area that they supply with CO\(_2\). To understand why
stomata are not ideally dispersed, more modelling should be done to
estimate the fitness gain of stomatal dispersion. Additionally, genetic
manipulation studies should attempt to create mutants with clustered and
ideally dispersed stomata for a comparison of their photosynthetic
traits. This could have extremely important implications for maximum
assimilation rates in crops as most crop species are grown in high light
where CO\(_2\) is often limiting. In drought-prone environments,
increased stomatal dispersion may increase water use efficiency by
reducing the number of stomata needed to achieve the same internal
CO\(_2\) concentration, \(C_\text{i}\).

Beyond disperson on a single surface, gas exchange can be optimized via
stomatal coordination of abaxial and adaxial surfaces in amphistomatous
leaves. Given that leaf thicknesses are generally multiple times greater
than interstomatal distance (GIVE DISTANCES HERE). As a result,
abaxial-adaxial stomatal coordination reduces CO\(_2\) diffusion path
length far less than single surface dispersion, so we hypothesize this
strategy to afford less photosynthetic advantage to the leaf. Our
modelling results demonstrate that, even in ideal conditions, i.e.~thick
leaf, low stomatal densities, high light, low leaf porosity, high
Rubisco concentration, etc., the photosynthetic advantage of
coordination is minimal. We are not surprised by these results, but
still highlight them here as we are the first to report this finding.

Amphistomy is a unique and important adaptation found around the world
across many plant lineages \citep{muir_light_2018}, yet much of the
dynamics of amphistomy remain poorly understood. Here, we show that in
\emph{Arabidopsis thaliana} 1) stomata in are non-randomly dispersed,
but not ideally dispersed; 2) stomatal size and density are modulated by
light; 3) stomatal size is positively correlated with the area to which
it supplies CO\(_2\); and 4) abaxial-adaxial stomatal coordination is
not exhibited and is not shown to provide a strong photosynthetic
advantage using CO\(_2\) diffusion models. Interestingly, these findings
did not validate many of our hypotheses which were based on first
principles, suggesting that there may be limits on plants' ability to
control stomatal placement. Future studies which elucidate these
limitations may have important implications for agriculturaly
productivity in a rapidly changing world.

@@

>>>>

\hypersetup{
  pdftitle={Does stomatal patterning in amphistomatous leaves minimize the CO\_2 diffusion path length with leaves of Arabidopsis thaliana?},
  pdfkeywords={amphistomy, Arabidopsis thaliana, CO\(_2\)
diffusion, finite element method, optimality, photosynthesis, stomata},
  breaklinks=true,
  bookmarks=true,
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

@@

% Create labels for Addresses if the are given in Elsevier format
   \MyNewLabel{ADDRTXT}{UHM}{%
  School of Life Sciences, University of Hawaiʻi at Mānoa, Honolulu, HI
96822%
 }
   \MyNewLabel{ADDRTXT}{CU}{%
  Ecology and Evolutionary Biology, Universtity of Colorado, Boulder, CO
80309%
 }
   \MyNewLabel{ADDRTXT}{NIAB}{%
  National Institute of Agricultural Botany, Cambridge, United Kingdom
CB3 0LE%
 }
   \MyNewLabel{ADDRTXT}{UCD}{%
  Department of Plant Sciences, University of California, Davis, CA
95616%
 }
   \MyNewLabel{ADDRTXT}{UWM}{%
  Department of Botany, University of Wisconsin, Madison, WI 53706%
 }

% Create labels for Footnotes if they are given in Elsevier format

% Pandoc header-include feature
\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{lineno}
\linenumbers
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}
\usepackage{caption}
\usepackage{newunicodechar,graphicx}
\DeclareRobustCommand{\okina}{\raisebox{\dimexpr\fontcharht\font`A-\height}{\scalebox{0.8}{`}}}
\newunicodechar{ʻ}{\okina}
% Pandoc header-include feature
\usepackage{booktabs}

@@

\title[]{Does stomatal patterning in amphistomatous leaves minimize the
CO\(_2\) diffusion path length with leaves of \emph{Arabidopsis
thaliana}?}

@@

\author[%
\addrlblbycode{UHM},\addrlblbycode{CU}%
,\refstepcounter{correspcnt}\setcounter{thisauthcorresp}{\value{correspcnt}}\fnsymbol{thisauthcorresp}%
%
%
]{Jacob L. Watts}

\addrbycode{UHM}
\addrbycode{CU}

\corresp[\fnsymbol{thisauthcorresp}]{Corresponding author. \href{mailto:Jacob.Watts-1@colorado.edu}{\nolinkurl{Jacob.Watts-1@colorado.edu}}}

\author[%
\addrlblbycode{NIAB}%
%
%
%
]{Graham Dow}

\addrbycode{NIAB}

\author[%
\addrlblbycode{UCD}%
%
%
%
]{Thomas N. Buckley}

\addrbycode{UCD}

\author[%
\addrlblbycode{UHM},\addrlblbycode{UWM}%
%
%
%
]{Christopher D. Muir}

\addrbycode{UHM}
\addrbycode{UWM}

@@

\abstract{
We will write abstract after Discussion and Conclusions are complete}

@@

Many successful predictions about stomata and other leaf traits can be
made by hypothesizing that natural selection should optimizes CO\(_2\)
gain per unit of water loss
\citep{cowan_stomatal_1977, buckley_optimal_2017, sperry_predicting_2017}.
However, stomatal anatomy may be partially constrained by physical and
developments limits on phenotypic expression
\citep{croxdale_stomatal_2000, harrison_influence_2020, muir_how_2023}.
Sometimes optimization leads to similar phenotypes across many disparate
species. For example, almost all stomata follow the one cell spacing
rule to maintain proper stomatal functioning
\citep{geisler_oriented_2000, dow_physiological_2014}; however some
species (notably in \emph{Begonia}) appear to benefit from overlapping
vapor shells caused by stomatal clustering
\citep{yi_gan_stomatal_2010, lehmann_effects_2015, papanatsiou_stomatal_2017}.
Stomatal traits also vary adaptively in different environments. Stomatal
density positively co-varies with irradiance during leaf development and
negatively co-varies with CO\(_2\) concentration
\citep{gay_influence_1975, schoch_dependence_1980, woodward_stomatal_1987, royer_stomatal_2001},
consistent with optimality predictions. Stomatal size is jointly
controlled by genome size, light, and stomatal density
\citep{jordan_environmental_2015}. Size positively co-varies with genome
size \citep{roddy_scaling_2020} and negatively co-varies with stomatal
density \citep{camargo_density_2011}. Total stomatal area (size
\(\times\) density) is optimized for operational conductance
(\(g_\text{s,op}\)) rather than maximum conductance (\(g_\text{s,max}\))
such that stomatal apertures are most responsive to changes in the
environment at their operational aperture
\citep{franks_physiological_2012, liu_scaling_2021}. Stomatal aperture
can compensate for maladaptive stomatal densities to an extent
\citep{bussis_stomatal_2006}, but stomatal density and size ultimately
determine a leaf's theoretical \(g_\text{s,max}\)
\citep{sack_developmental_2016}, which is proportional to
\(g_\text{s,op}\) \citep{murray_consistent_2020}. Additionally, low
stomatal densities lead to irregular and insufficient CO\(_2\) supply
and reduced photosynthetic efficiency in areas far from stomata
\citep{pieruschka_lateral_2006, morison_lateral_2005}, while high
stomatal densities can reduce water use efficiency (WUE)
\citep{bussis_stomatal_2006} and incur excessive metabolic costs
\citep{deans_optimization_2020}. In most species, stomata occur on the
abaxial (usually lower) leaf surface; but amphistomy, the occurrence of
stomata on both abaxial and adaxial leaf surfaces, is also prevalent in
high light environments with constant or intermittent access to
sufficient water
\citep{mott_adaptive_1982, jordan_using_2014, muir_light_2018, drake_two_2019, muir_is_2019}.
Amphistomy effectively halves the CO\(_2\) diffusion path length and
boundary layer resistance by doubling boundary layer conductance
\citep{parkhurst_adaptive_1978, harrison_influence_2020, mott_amphistomy_1991}.
Historically, stomatal patterning in dicot angiosperms was thought to be
random with an exclusionary distance surrounding each stomate
\citep{sachs_developmental_1974}; however, the developmental controls of
stomatal patterning are poorly understood and likely more complex than
random development along the leaf surface.
\citet{croxdale_stomatal_2000}{]} reviews three developmental theories
which attempt to explain stomatal patterning in angiosperms: inhibition,
cell lineage, and cell cycle, ultimately arguing for a cell cycle based
control of stomatal patterning.

@@

We hypothesized that natural selection will favor stomatal patterning
and distribution to minimize the diffusion path length. In
amphistomatous leaves, this would be accomplished by 1) a dispersed,
equilateral triangular distribution of stomata on both abaxial and
adaxial leaf surfaces and 2) coordinated stomatal spacing on each
surface that offsets the position of stomata
(\autoref{fig:ideal-amphi-grid}). Coordination between leaf surfaces is
defined in this study as the occurrence of stomata in areas farthest
from stomata on the opposite leaf surface. Additionally, because
CO\(_2\) is more limiting for photosynthesis under high light, we
hypothesize that in high light 3) there should be more stomata, and 4)
stomata should be more uniformly distributed than in low light. Finally,
as stomatal densities are selected for optimal operational aperture, we
hypothesize that 5) stomatal length will be positively correlated with
the area of the leaf surface to which it is closest. We refer to this as
the `stomatal zone', the leaf area surrounding a focal stomate closest
to that stomate and therefore the zone it supplies with CO\(_2\)). This
way, each stomate can be optimally sized relative to the mesophyll
volume it supplies.

\begin{figure}[ht]
\includegraphics[width = 4in]{figures/ideal-amphi-grid.pdf}
\caption{Idealized amphistomatous stomatal grid with uniform stomatal patterning and perfect abaxial-adaxial coordination.}
\label{fig:ideal-amphi-grid}
\end{figure}

@@

\begin{table}[ht]
\centering
\begin{tabular}{ll}
  \hline
trait & relationship \\ 
  \hline
leaf thickness & + \\ 
  interstomatal distance & + \\ 
  leaf porosity & - \\ 
  light & + \\ 
   \hline
\end{tabular}
\caption{A summary of the hypothesized relationships between leaf traits and environmental conditions and photosynthetic advantage of stomatal spatial coordination in amphistomatous leaves.} 
\label{tab:hypotheses}
\end{table}

@@

{[}Graham - can you fill in details on growth conditions, image
aquisitiion, and image dimensions?{]}

@@

For each sample image, we also simulated \(10^3\) synthetic data with
\(n\) stomata ideally dispersed in an equilateral triangular grid. For
these grids, we integrated over plausible stomatal densities and then
conditioned on stomatal grids with exactly \(n\) stomata. The simulated
stomatal count was drawn from a Poisson distribution with the mean
parameter \(\lambda\) drawn from a Gamma distribution with shape \(n\)
and scale 1 (\(\lambda \sim \Gamma(n, 1)\)). \(\Gamma(n, 1)\) is the
posterior distribution of \(\lambda\) with a flat prior distribution.
This allows us to integrate over uncertainty in the stomatal density
from the sample image.

@@

\begin{figure}[ht]
\includegraphics[width = 5in]{figures/model_summary.pdf}
\caption{There is little photosynthetic benefit of offsetting stomatal position each surface based on a 2-D model of photosynthesis. The coordination advantage (\autoref{eq:coordination_advantage}) is close to zero under nearly all of the parameter space \autoref{tab:model_var}, meaning that the photosynthetic rate of amphistomatous leaves with stomata optimally offset is nearly equal to leaves with stomata on each surface in the same position along the leaf plane. $I_0$: PPFD incident on the leaf surface; $\varphi_\text{pal}$: Fraction of intercellular airspace (aka porosity), palisade; $T_\text{leaf}$: Leaf thickness; $U$: Interstomatal distance.}
\label{fig:model_summary}
\end{figure}

@@

Stomata cost resources to maintain \citep{deans_optimization_2020} and
expose leaves to risks such as hydraulic failure
\citep{wang_theoretical_2020} or infection by plant pathogens
\citep{melotto_stomatal_2017}. Therefore leaves should develop enough
stomata to adequately supply CO\(_2\) to chloroplasts, but not
overinvest. A widespread hypothesis in plant ecophysiology is that
natural selection optimizes traits like stomatal size, density, and
distribution to maximize carbon gain relative to any costs in a given
environmental context. In principle, spacing stomata to minimize the
average distance between stomatal pores and chloroplasts within the
mesophyll should increase carbon gain, all else being equal. However,
reducing this distance to its absolute minimum may be constrained by
developmental processes, or the photosynthetic benefit may be too small
to be `seen' by natural selection (i.e.~the selection coefficient is
less than drift barrier \citep{sung_drift-barrier_2012}).

We tested five related hypotheses about stomatal spacing in
amphistomatous leaves using the model angiosperm \emph{Arabidopsis
thaliana} grown under different light intensities. First, we predicted
that stomata on each surface are overdispersed relative to a random
uniform distribution, which should increase CO\(_2\) supply. Stomata on
each surface are overdispersed (\autoref{fig:single-surface}), but are
not ideally dispersed in an equilateral triangular grid as would be
optimal to minimize CO\(_2\) diffusion path length and equalize the area
supplied by each stomate (\autoref{fig:tessellation-example}). Second,
we predicted that an optimal amphistomatous leaf has offset stomata such
that stomata are more likely to appear on one leaf surface if there is
not a stomata directly opposite it on the other surface as shown in
\autoref{fig:ideal-amphi-grid}. However, there is no evidence for
coordination and the positions on each surface appear independent,
regardless of light treatment (\autoref{fig:dual-surface}). Third, we
predicted that plants respond plastically to higher light intensity by
increasing stomatal density. \emph{Arabidopsis} plants grown under high
light had higher stomatal density than the same genotype growns under
low and medium light intensity (\autoref{fig:density}). However, we
found no support for our fourth prediction that stomata would be more
evenly dispersed at high light intensity (\autoref{fig:single-surface}).
Finally, we predicted that within leaf variation in stomatal size would
correlate with stomatal spacing, as larger stomata can supply larger
volumes of adjacent mesophyll. In all three light treatments, stomatal
size positively covaried with the stomatal zone, i.e.~adjacent region of
mesophyll that would be supplied by that stomate
(\autoref{fig:length-area}).

Stomatal spacing on \emph{A. thaliana} leaves partially supports our
overall hypothesis that natural selection minimizes the average distance
between stomata and chloroplasts, for a given overall stomatal density.
There are three nonmutually exclusive hypotheses for why several of our
predictions were wrong. First, our predictions are wrong becuase they
are based on overly simplistic assumptions about epidermal and mesophyll
anatomy. Second, natural selection may be constrained by developmental
processes that prevent phenotypes from reaching their adaptive optima.
Third, the benefit of some traits may be of too little consequence to
result in fitness differences large enough to be respond to selection.
We consider the plausibility of these alternative hypotheses below and
present ideas for future work to test them.

{[}this paragraph should discuss why our assumptions may be too
simplistic{]} We assume an idealized leaf epidermal and mesophyll
structure that is homogenous and unconstrained by other tradeoffs. Real
leaves not only provide pathways for CO\(_2\) diffusion, but must supply
water, intercept light, and deter herbivores and pathogens. These
competing interests results in nonuniform epidermal and mesophyll
structure that could alter our predictions about optimal stomatal
spacing. {[}explain some examples{]}

{[}this paragraph should discuss developmental constraint{]} No
developmental pathway exists to ensure the ideal placement of stomata on
the leaf. {[}add more explanation{]}

{[}this paragraph should discuss limits on natural selection{]} The gas
exchange model demonstrate little photosynthetic gain from
abaxial-adaxial stomatal coordination (\autoref{fig:model_summary}).
Even though lateral diffusion may limit photosynthesis
\citep{morison_lateral_2005}, the marginal gain from optimally
offsetting stomata is not sufficient to generate fitness differences
relative to the strength of genetic drift (i.e.~the drift-barrier). We
can similarly extrapolate that an ideal, equilateral triangular stomatal
spacing is only slightly better than a suboptimal overdispered pattern.
Explaining these observations as the result of weak selection is in
tension with the finding that stomatal size and zone positively covary,
which would suggest that small changes in lateral diffusion distance are
significant. As described above, the positive correlation between
stomatal size and zone may be explained by common developmental
processes rather than as an adaptation to maximize CO\(_2\) diffusion.

{[}discuss significance of our study and future directions{]} Our study
corroborates previous studies which demonstrate that stomata are
non-randomly distributed along the leaf surface as a result of
developmental mechanisms such as spatially biased arrest of stomatal
initials \citep{boetsch_arrest_1995}, oriented asymmetric cell division
\citep{geisler_oriented_2000}, and cell cycle controls
\citep{croxdale_stomatal_2000}. We do not investigate the potential
developmental pathways that influence stomatal dispersion in this study;
however, they are important to consider as these pathways could limit
plants from reaching the theoretical peak in the adaptive landscape:
uniform stomatal dispersion. Instead, as this study suggests, plants may
simply compensate with higher stomatal density and by fitting stomatal
size to the area that they supply with CO\(_2\). To understand why
stomata are not ideally dispersed, more modelling should be done to
estimate the fitness gain of stomatal dispersion. Additionally, genetic
manipulation studies should attempt to create mutants with clustered and
ideally dispersed stomata for a comparison of their photosynthetic
traits. This could have extremely important implications for maximum
assimilation rates in crops as most crop species are grown in high light
where CO\(_2\) is often limiting. In drought-prone environments,
increased stomatal dispersion may increase water use efficiency by
reducing the number of stomata needed to achieve the same internal
CO\(_2\) concentration, \(C_\text{i}\).

{[}Conclusion{]} Our results suggest that after optimiziing stomatal
density and having developmental rules for spacing stomata relatively
evenly, there may be limited gains to further optimization. Therefore,
developmental constraints may be necessary to make sense of some
features of stomatal spacing and distribution.

@@

\section{Competing interests}

The authors declare no competing interests.

\section{Author contributions statement}

JLW and CDM conceived of the project, analyzed data, and wrote the
manuscript. GW provided data. TNB contributed to model development and
helped edit the manuscript.
